

\documentclass{article}
\usepackage{fullpage, amsmath}
\pagestyle{empty}

\begin{document}

\textwidth = 6.5in
\textheight = 9.7in
\oddsidemargin = 0.0in
\evensidemargin = 0.0in
\topmargin = 0.0in
\headheight = -0.45in
\headsep = 0.0in
\parskip = 0.2in
\parindent = 0.0in

\pagestyle{empty}


\centerline{Lab 4}

\noindent\rule{6.7in}{0.01in}

\vspace{0.2in}

The \textit{binomial theorem} gives a way to expand out an expression of the form $(a+b)^n$:
\[(a+b)^n=\sum_{k=0}^n\binom{n}{k}a^{n-k}b^k.\]

For example, if we wanted to expand the polynomial $(x+2)^3$ we could apply the binomial theorem and find that
\begin{align*}
(x+2)^3 &= \binom{3}{0}x^32^0+\binom{3}{1}x^22^1+\binom{3}{2}x^12^2+\binom{3}{3}x^02^3 \\
&= 1x^32^0+3x^22^1+3x^12^2+1x^02^3 \\
&= x^3+6x^2+12x+8. \\
\end{align*}


Use the binomial theorem to answer the following questions. This lab is broken into a group portion (worth 4 points) and an individual portion (worth 6 points).


\noindent\textbf{Group Questions:} Hand the solutions in for these questions by the end of the lab period.
\begin{enumerate}

\item (1/2 point) Expand $(x+y)^5$. What is the degree of this polynomial?

\vspace{0.2in}

\item (1/2 point) What is the coefficient of $x^{101}y^{99}$ in the expansion of $(2x-3y)^{200}?$

\vspace{0.2in}

\item (1 point) Give a formula for the coefficient of $x^k$ in the expansion of $(x+1/x)^{100}$, where $k$ is an integer.

\vspace{0.2in}

\item (2 points) Prove the binomial theorem using mathematical induction. 

\end{enumerate}

\vspace{0.5in} 

\noindent\textbf{Individual Questions:} On back. Due in two weeks.

\newpage

\noindent\textbf{Individual Questions:} 

\begin{enumerate}

\item (2 points)  A polynomial in any number of variables is called \textit{homogeneous} if the sums of the exponents on the variables in each term add up to the same number. For example, $x^2+2xy+y^2$ is homogeneous because the exponents on $x$ and $y$ in each term add to 2. For which values of $k$ is $(x+y)^k$ homogeneous? Prove your answer (using full sentences!).

\vspace{0.2in}

\item (1 point) For what values of $k$, if any, is $(x+y+z)^k$ a homogeneous polynomial? When the polynomial is homogeneous, what is the degree of the polynomial? You do not have to prove your answer, just state it.

\vspace{0.2in}

\item (3 points) What is the coefficient of $w^4x^2yz$ in the expansion of $(w+2x+y-z)^8$? You do not need to write a proof, but please explain your answer well.

\end{enumerate}

\end{document}